Higher-Order unification for free!

### 2024-04-29 paper/main.tex
update tex
remove 2 todos

\begin{proposition}[Good unification]\label{prop:good-unif}
A good unification $\UnifRel\;$ for equality $\;\EqualRel\;$ in the domain
$\mathcal{D}$ satisfies the following properties:\noteD{use $\sigma$, not $\rho$}
  \begin{gather}
    \unifcorrectX{\EqualRel}{\UnifRel}{t}{\{~t_1,~ t_2~\} \subseteq \mathcal{D}}\\
    \unifcompleteX{\EqualRel}{\UnifRel}{t}{\{~ t_1,~ t_2 ~\} \subseteq \mathcal{D}}
  \end{gather}
\end{proposition}

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms\noteD{module var reflexivity}. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
<<\elpiIn{fuva X ~\Uo~s}>>. In this case, after compilation, we have
<<\elpiIn{uva Y [] ~\Ue~t}>>
that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compile to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \ref{clause:lam-lam}
          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
          to unify, it cannot be because of the $\lambda$ constructor in the head of $t_1$,
          since \ref{clause:lam-lam} or \ref{clause:eta1} try to unify under it, eventually finding two
          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.

>>>>

\begin{proposition}[Good unification]\label{prop:good-unif}
A good unification $\UnifRel\;$ for equality $\;\EqualRel\;$ in the domain
$\mathcal{D}$ satisfies the following properties:%\noteD{use $\sigma$, not $\rho$}
  \begin{gather}
    \unifcorrectX{\EqualRel}{\UnifRel}{t}{\{~t_1,~ t_2~\} \subseteq \mathcal{D}}\\
    \unifcompleteX{\EqualRel}{\UnifRel}{t}{\{~ t_1,~ t_2 ~\} \subseteq \mathcal{D}}
  \end{gather}
\end{proposition}

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
In this setting, \Ee is as strong as \Eo on ground terms and reflexivity for
unification variables. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
<<\elpiIn{fuva X ~\Uo~s}>>. In this case, after compilation, we have
<<\elpiIn{uva Y [] ~\Ue~t}>>
that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{proof}[Proof sketch] %\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compile to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \ref{clause:lam-lam}
          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
          to unify, it cannot be because of the $\lambda$ constructor in the head of $t_1$,
          since \ref{clause:lam-lam} or \ref{clause:eta1} try to unify under it, eventually finding two
          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.
